---
title: "Occ Health Notes"
author: "Chris Austin"
date: "2023-08-02"
output: html_document

---

```{r include=FALSE}
library(ggplot2)
library(plotly)

knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
<br>

## Employer-reported respiratory illnesses increased nearly 3,000 percent in 2020
<br>
The Survey of Occupational Injuries and Illnesses (SOII) is a Federal/State cooperative program that publishes estimates on nonfatal occupational injuries and illnesses. Each year, approximately 200,000 employers report for establishments in private industry and the public sector (state and local government). In-scope cases include work-related injuries or illnesses to workers who require medical care beyond first aid. 

The SOII relies on Occupational Safety and Health Administration (OSHA) recordkeeping requirements, which mandate employers record certain work-related injuries and illnesses. COVID-19 is considered a respiratory illness under [criteria](https://www.osha.gov/coronavirus/standards) established by OSHA. The **illness is recordable if a worker is infected while performing work-related duties**. Once collected for the SOII, cases requiring at least one day away from work are coded using the [Occupational Injury and Illness Classification System (OIICS)](https://www.bls.gov/iif/definitions/occupational-injuries-and-illnesses-classification-manual.htm).
<br>
<br>
<br>

### Total reported injury and illness case counts, private industry, Nebraska 2019-2020

```{r reportabe_cases, echo=FALSE, warning=FALSE}

reportabe_cases <- tibble::data_frame(SeriesID  = c("Total reported cases", "Injuries", "Illness", "Total reported cases", "Injuries", "Illness"),
                  year = c("2019", "2019", "2019", "2020", "2020", "2020"),
                  cases = c(20400, 19000, 1300, 19300, 15200, 4100)
                  )

reportabe_cases$SeriesID <- factor(reportabe_cases$SeriesID, levels = c("Total reported cases", "Injuries", "Illness"))


p <- ggplot(reportabe_cases, aes(SeriesID, cases, fill=year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  labs(x = NULL, y = "Cases", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")

ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))
```
<br>
<br>

### Counts of cases of all illnesses and respiratory illnesses, private industry, Nebraska, 2016-2020

```{r illnesscases, echo=FALSE, warning=FALSE}

illness_cases <- tibble::data_frame(SeriesID = c("All illnesses",	"All illnesses",	"All illnesses",	"All illnesses",	"All illnesses",	"Respiratory illnesses",	"Respiratory illnesses",	"Respiratory illnesses",	"Respiratory illnesses",	"Respiratory illnesses"), 
year = c("2016",	"2017",	"2018",	"2019",	"2020",	"2016",	"2017",	"2018",	"2019",	"2020"),
Cases = c(2000,	1700,	1600,	1300,	4100,	0,	0,	100,	100,	3200))


p <- ggplot(illness_cases, aes(year, Cases, fill=SeriesID)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  labs(x = NULL, y = "Cases", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))
```
* 78 percent of employer-reported illness were respiratory illnesses in 2020.
* Employer-reported respiratory illnesses increased 3,000 percent in 2020

<br>
<br>
<br>

## Respirtory illnesses in accross selected industries in Nebraska

<br>
Nationally, women accounted for <ins>71 percent</ins> of the 390,020 reported workplace illness cases caused by "other diseases due to viruses, not elsewhere classified" in private establishments in 2020.

In Nebraska, women accounted for <ins>65 percent</ins> of the 2,890 reported workplace illness cases caused by "other diseases due to viruses, not elsewhere classified" in private establishments in 2020

* Women in Nebraska had **2.7 times** the rate of “other diseases due to viruses, not elsewhere classified” as compared to men.
```{r viruscases, echo=FALSE, warning=FALSE}

resp_cases <- tibble::data_frame(industry = c("Healthcare & Social Assistance", "Healthcare & Social Assistance", "Healthcare & Social Assistance", "Manufactoring", "Manufactoring", "Manufactoring", "Retail", "Retail", "Retail"),
Year = c("2019", "2020", "2021", "2019", "2020", "2021", "2019", "2020", "2021", "2019", "2020", "2021"), 
Rates = c(3.8,	160.9,	41.1,	2,	126.5,	59.9,	0,	30,	36.1))

resp_cases$industry <- factor(resp_cases$industry, levels = c("Healthcare & Social Assistance", "Manufactoring", "Retail"))

library(ggplot2)
library(plotly)

p <- ggplot(resp_cases, aes(industry, Rates, fill=Year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  labs(x = NULL, y = "Cases", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))

```
<br>
<br>


### Rate of workplace illnesses involving days away from work caused by other diseases due to viruses, n.e.c., by selected occupation, Nebraska, 2020

```{r occupaitonvirus, echo=FALSE, warning=FALSE}
occupation_virus <- tibble::data_frame(Occupation = c("Healthcare practitioners and technical",	"Production",	"Service",	"Transportation and material moving",	"Installation, maintenance, and repair",	"Sales and related",	"Management, business, financial",	"Educational instruction and library occupations",	"Office and administrative support",	"Computer, engineering, and science"),
Rates = c(127.9, 116.6, 79.9,	46.7,	29.8,	19.7,	17.8,	17.6,	8.4, 7.4))

occupation_virus$Occupation <- factor(occupation_virus$Occupation,              
                  levels = occupation_virus$Occupation[order(occupation_virus$Rates, decreasing = FALSE)])


library(ggplot2)
library(plotly)

p <- ggplot(occupation_virus, aes(y = Occupation, x = Rates, group = Occupation)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", fill='#0C475B', width = .7) + 
  labs(y = NULL, x = "Rate per 10,000 FTE", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_classic() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.25, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))
```


```{r, occ health sys, echo=FALSE, warning=FALSE}

suppressPackageStartupMessages({
  library(DBI)
  library(odbc)
  library(tidyverse)
  library(dbplyr)
  library(tidytext)
})

####              create connection                  ####
#########################################################

con <- dbConnect(odbc(),
                 Driver = "SQL Server",
                 Server = "dhhsessencesql1.stone.ne.gov",
                 Database = "NE_ESSENCE5_Detection", 
                 UID    ="<REDACTED_USER>",
                 PWD    ="<REDACTED_PASSWORD>",
                 port = 1433)


con2 <- dbConnect(odbc(),
                  Driver = "SQL Server",
                  Server = "NEDSSQLAG2.stone.ne.gov",
                  Database = "ER_SURVEILLANCE_PROD", 
                  UID    ="<REDACTED_USER>",
                  PWD    ="<REDACTED_PASSWORD>",
                  port = 1433)

```

```{r, warning=FALSE}
################################################################################

 #                           query database                                 #

################################################################################

reportdate <- as.Date(Sys.Date())

ED_data_df <- tbl(con, "View_Cache_ER_Base") |>
  select(Date, MonthYear, Region, Age, Sex, ChiefComplaintParsed, DischargeDiagnosis, TriageNotesOrig, DischargeDisposition, HasBeenE, PatientSrcOfPayment) |>
  filter(between(Date, as.Date("2019-01-01"), reportdate) & HasBeenE == 1 & Region != "OTHER_REGION") |>
  left_join(tbl(con,"DispositionCategoryMappingTable"), by = c("DischargeDisposition" = "Disposition")) |>
  mutate(Age_grp = case_when(between(Age, 0, 10) ~ "0-10",
                             between(Age, 11, 14) ~ "11-14",
                             between(Age, 15, 24) ~ "15-24",
                             between(Age, 25, 34) ~ "25-34",
                             between(Age, 35, 44) ~ "35-44",
                             between(Age, 45, 54) ~ "45-54",
                             between(Age, 55, 64) ~ "55-64",
                             Age >= 65 ~ "65+")) |>
  arrange(desc(Date)) |>
  collect() |>
  filter(str_detect(DischargeDiagnosis, regex('Y99.0|Y990|Y99.1|Y991|Y92.61|Y9261|Y92.62|Y9262|Y92.63|Y9263|Y92.64|Y9264|Y92.65|Y9265|Y92.69|Y9269|Y92.71|Y9271|Y92.72|Y9272|Y92.73|Y9273|Y92.74|Y9274|Y92.79|Y9279|Z04.2|Z042|Z57.6|Z576|Z57.8|Z578', ignore_case = TRUE)) |
  str_detect(TriageNotesOrig, regex("\\bat work\\b|\\bwork related\\b|\\bwork acc\\b|\\bwork inj\\b|\\bworkers comp\\b|\\bworker comp\\b|\\bworkmans comp\\b|\\bworkmens comp\\b|\\baccident work\\b|\\bjob related\\b|\\bon the job\\b|\\bOJI\\b|\\bOTJ\\b", ignore_case = TRUE)) & 
  str_detect(TriageNotesOrig, regex("workout|\\bworking out\\b|\\bwork up\\b|\\bwork note\\b|labwork|\\blab work\\b|bloodwork|\\bblood work\\b", ignore_case = TRUE), negate = TRUE)
  )
  

ED_data_df$year = lubridate::year(ED_data_df$Date)

occ_related <- ED_data_df |>
  mutate(event_of_exposure = case_when(
    str_detect(ChiefComplaintParsed, regex("fall|slip|trip", ignore_case = TRUE)) ~ "Fall Slips Trips", 
    str_detect(ChiefComplaintParsed, regex("assault|assaulted", ignore_case = TRUE)) ~ "Assault ", 
    str_detect(ChiefComplaintParsed, regex("mva|vehicle|truck|tractor|combine|collision|crashed", ignore_case = TRUE)) ~ "MVA",
    str_detect(ChiefComplaintParsed, regex("fire|explosion", ignore_case = TRUE)) ~ "Fire or Explosion", 
    str_detect(ChiefComplaintParsed, regex("bite|bit by", ignore_case = TRUE)) ~ "Bites",
    str_detect(ChiefComplaintParsed, regex("struck by", ignore_case = TRUE)) ~ "Struck By",
    str_detect(ChiefComplaintParsed, regex("forklift", ignore_case = TRUE)) ~ "Forklift")) |>
  mutate(part_of_body = case_when(
    str_detect(ChiefComplaintParsed, regex("eye", ignore_case = TRUE)) ~ "Eye",
    str_detect(ChiefComplaintParsed, regex("head", ignore_case = TRUE)) ~ "Head", 
    str_detect(ChiefComplaintParsed, regex("neck", ignore_case = TRUE)) ~ "Neck",
    str_detect(ChiefComplaintParsed, regex("back", ignore_case = TRUE)) ~ "Back", 
    str_detect(ChiefComplaintParsed, regex("shoulder", ignore_case = TRUE)) ~ "Shoulder", 
    str_detect(ChiefComplaintParsed, regex("leg", ignore_case = TRUE)) ~ "Leg", 
    str_detect(ChiefComplaintParsed, regex("knee", ignore_case = TRUE)) ~ "Knee", 
    str_detect(ChiefComplaintParsed, regex("toe|foot", ignore_case = TRUE)) ~ "Foot",
    str_detect(ChiefComplaintParsed, regex("arm", ignore_case = TRUE)) ~ "Arm",
    str_detect(ChiefComplaintParsed, regex("wrist", ignore_case = TRUE)) ~ "Wrist",
    str_detect(ChiefComplaintParsed, regex("finger|hand|thumb", ignore_case = TRUE)) ~ "Hand")) |>
  mutate(nature_of_injury = case_when(
    str_detect(ChiefComplaintParsed, regex("cut|laceration", ignore_case = TRUE)) ~ "Cut/Laceration", 
    str_detect(ChiefComplaintParsed, regex("burn", ignore_case = TRUE)) ~ "Burns", 
    str_detect(ChiefComplaintParsed, regex("contusion", ignore_case = TRUE)) ~ "Contusion", 
    str_detect(ChiefComplaintParsed, regex("sprain|strains|tears", ignore_case = TRUE)) ~ "Sprain Strains Tears", 
    str_detect(ChiefComplaintParsed, regex("fracture", ignore_case = TRUE)) ~ "Fracture",
    str_detect(ChiefComplaintParsed, regex("amputation", ignore_case = TRUE)) ~ "Amputation"))

occ_related_freq_event <- occ_related |>
  group_by(event_of_exposure, year) |>
  summarise(n = n()) |>
  na.omit()


occ_related_freq_event$year <- as.character(occ_related_freq_event$year)


occ_event <- ggplot(data = occ_related_freq_event, aes(x = event_of_exposure, y = n, fill = year, group = year)) +
  geom_bar(color = "black", position="dodge", stat="identity") +
  scale_fill_brewer(palette = "Set2") + 
  theme_bw()


occ_related_freq_body <- occ_related |>
  group_by(part_of_body, year) |>
  summarise(n = n()) |>
  na.omit()

occ_related_freq_body$year <- as.character(occ_related_freq_body$year)


occ_body <- ggplot(data = occ_related_freq_body, aes(x = part_of_body, y = n, fill = year, group = year)) +
  geom_bar(color = "black", position="dodge", stat="identity") +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() 
  

occ_related_freq_nature <- occ_related |>
  group_by(nature_of_injury, year) |>
  summarise(n = n()) |>
  na.omit()

occ_related_freq_nature$year <- as.character(occ_related_freq_nature$year)


occ_nature <- ggplot(data = occ_related_freq_nature, aes(x = nature_of_injury, y = n, fill = year, group = year)) +
  geom_bar(color = "black", position="dodge", stat="identity") +
  scale_fill_brewer(palette = "Set2") +
  theme_bw()
 
  
occ_related_freq_age <- ED_data_df |>
  group_by(Age_grp, year) |>
  summarise(n = n()) |>
  na.omit()

occ_related_freq_age$year <- as.character(occ_related_freq_age$year)


occ_age <- ggplot(data = occ_related_freq_age, aes(x = Age_grp, y = n, fill = year, group = year)) +
  geom_bar(color = "black", position="dodge", stat="identity") +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  facet_wrap(~ Age_grp, scales = "free_y")
```

## Occupational Health Syndromic Surveillance Preview

<br>
<br>

### Injuries by Event
```{r, occ health plots, echo=FALSE, warning=FALSE}

ggplotly(occ_event) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.25, text = "Source: ESSENCE",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)))

```

<br>
<br>

### Injuries to Body
```{r, occ health plots 2, echo=FALSE, warning=FALSE}

ggplotly(occ_body) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.25, text = "Source: ESSENCE",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)))

```

<br>
<br>

### Nature of Injuries
```{r, occ health plots 3, echo=FALSE, warning=FALSE}

ggplotly(occ_nature) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.25, text = "Source: ESSENCE",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)))

```
