---
title: "Occ health employer-reported respiratory illnesses report"
author: "Chris Austin"
date: "2023-08-02"
output: html_document

---

```{r include=FALSE}
library(ggplot2)
library(plotly)

knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
<br>

# Employer-reported respiratory illnesses increased nearly 3,000 percent in 2020
<br>
The Survey of Occupational Injuries and Illnesses (SOII) is a Federal/State cooperative program that publishes estimates on nonfatal occupational injuries and illnesses. Each year, approximately 200,000 employers report for establishments in private industry and the public sector (state and local government). In-scope cases include work-related injuries or illnesses to workers who require medical care beyond first aid. 

The SOII relies on Occupational Safety and Health Administration (OSHA) recordkeeping requirements, which mandate employers record certain work-related injuries and illnesses. COVID-19 is considered a respiratory illness under [criteria](https://www.osha.gov/coronavirus/standards) established by OSHA. The **illness is recordable if a worker is infected while performing work-related duties**. Once collected for the SOII, cases requiring at least one day away from work are coded using the [Occupational Injury and Illness Classification System (OIICS)](https://www.bls.gov/iif/definitions/occupational-injuries-and-illnesses-classification-manual.htm).
<br>
<br>
<br>

### Total reported injury and illness case counts, private industry, Nebraska 2019-2020

```{r reportabe_cases, echo=FALSE, warning=FALSE}

reportabe_cases <- tibble::data_frame(SeriesID  = c("Total reported cases", "Injuries", "Illness", "Total reported cases", "Injuries", "Illness"),
                  year = c("2019", "2019", "2019", "2020", "2020", "2020"),
                  cases = c(20400, 19000, 1300, 19300, 15200, 4100)
                  )

reportabe_cases$SeriesID <- factor(reportabe_cases$SeriesID, levels = c("Total reported cases", "Injuries", "Illness"))


p <- ggplot(reportabe_cases, aes(SeriesID, cases, fill=year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  scale_fill_manual(values=c("#f08080", "#ffdead")) +
  labs(x = NULL, y = "Cases", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")

ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))
```
<br>
<br>

### Cases of all illnesses and respiratory illnesses, private industry, Nebraska, 2016-20

```{r illnesscases, echo=FALSE, warning=FALSE}

illness_cases <- tibble::data_frame(SeriesID = c("All illnesses",	"All illnesses",	"All illnesses",	"All illnesses",	"All illnesses",	"Respiratory illnesses",	"Respiratory illnesses",	"Respiratory illnesses",	"Respiratory illnesses",	"Respiratory illnesses"), 
year = c("2016",	"2017",	"2018",	"2019",	"2020",	"2016",	"2017",	"2018",	"2019",	"2020"),
Cases = c(2000,	1700,	1600,	1300,	4100,	0,	0,	100,	100,	3200))


p <- ggplot(illness_cases, aes(year, Cases, fill=SeriesID)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  labs(x = NULL, y = "Cases", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))
```
* 78 percent of employer-reported illness were respiratory illnesses in 2020.
* Employer-reported respiratory illnesses increased 3,000 percent in 2020

<br>
<br>
<br>

### Rates of respiratory illnesses across selected industries in Nebraska, 2019-21

```{r indrates, echo=FALSE, warning=FALSE}

resp_cases <- tibble::data_frame(industry = c("Healthcare & Social Assistance", "Healthcare & Social Assistance", "Healthcare & Social Assistance", "Manufacturing", "Manufacturing", "Manufacturing", "Retail", "Retail", "Retail"),
Year = c("2019", "2020", "2021", "2019", "2020", "2021", "2019", "2020", "2021"), 
Rates = c(3.8,	160.9,	41.1,	2,	126.5,	59.9,	0,	30,	36.1))

resp_cases$industry <- factor(resp_cases$industry, levels = c("Healthcare & Social Assistance", "Manufacturing", "Retail"))

library(ggplot2)
library(plotly)

p <- ggplot(resp_cases, aes(industry, Rates, fill=Year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  scale_fill_manual(values=c("#f08080", "#ffdead", "#bc8f8f")) +
  labs(x = NULL, y = "per 10,000 FTE", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))

```
<br>
<br>


### Rate of respiratory illness per 10,000 full-time workers in healthcare & social assistance and nursing & residential care facilities, 2019-21 
<br>
In 2020, the respiratory illness rate for the healthcare & social assistance sector was **160.9 cases per 10,000 full-time workers**. That was notably higher than the rate of 3.8 before the COVID-19 pandemic in 2019. Within healthcare & social assistance, nursing & residential care facilities had the highest incidence rate of respiratory illness in 2020, **386.4 cases per 10,000 full-time workers**.

```{r HSArate, echo=FALSE, warning=FALSE}

resp_cases <- tibble::data_frame(industry = c("Healthcare & Social Assistance", "Healthcare & Social Assistance", "Healthcare & Social Assistance", "Nursing & residential care facilities", "Nursing & residential care facilities", "Nursing & residential care facilities"),
Year = c("2019", "2020", "2021", "2019", "2020", "2021"), 
Rates = c(3.8,	160.9,	41.1, 0,	386.4, 87.4))

resp_cases$industry <- factor(resp_cases$industry, levels = c("Healthcare & Social Assistance", "Nursing & residential care facilities"))

library(ggplot2)
library(plotly)

p <- ggplot(resp_cases, aes(industry, Rates, fill=Year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  scale_fill_manual(values=c("#f08080", "#ffdead", "#bc8f8f")) +
  labs(x = NULL, y = "per 10,000 FTE", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))


```


### Rate of respiratory illness per 10,000 full-time workers in manufacturing and animal slaughtering & processing, 2019-21 
<br>
In 2020, the respiratory illness rate for the manufacturing sector was **126.5 cases per 10,000 full-time workers**. That was notably higher than the rate of 2.0 before the COVID-19 pandemic in 2019. Within manufacturing, animal slaughtering & processing had the highest incidence rate of respiratory illness in 2020, **396.5 cases per 10,000 full-time workers**.


```{r MFGrate, echo=FALSE, warning=FALSE}

resp_cases <- tibble::data_frame(industry = c("Manufacturing", "Manufacturing", "Manufacturing", "Animal slaughtering & processing", "Animal slaughtering & processing", "Animal slaughtering & processing"),
Year = c("2019", "2020", "2021", "2019", "2020", "2021"), 
Rates = c(2,	126.5,	59.9,	0,	396.5,	180.9))

resp_cases$industry <- factor(resp_cases$industry, levels = c("Manufacturing", "Animal slaughtering & processing"))

library(ggplot2)
library(plotly)

p <- ggplot(resp_cases, aes(industry, Rates, fill=Year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  scale_fill_manual(values=c("#f08080", "#ffdead", "#bc8f8f")) +
  labs(x = NULL, y = "per 10,000 FTE", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))

```

### Rate of respiratory illness per 10,000 full-time workers in retail and general merchandise stores, 2019-21 
<br>
In 2021, the respiratory illness rate for the retail trade sector was **36.1 cases per 10,000 full-time workers**. That was notably higher before the COVID-19 pandemic in 2019. Within retail trade, general merchandise stores had the highest incidence rate of respiratory illness in 2021, **78.6 cases per 10,000 full-time workers**.

```{r RETrate, echo=FALSE, warning=FALSE}

resp_cases <- tibble::data_frame(industry = c("Retail", "Retail", "Retail", "General merchandise stores", "General merchandise stores", "General merchandise stores"),
Year = c("2019", "2020", "2021", "2019", "2020", "2021"), 
Rates = c(0, 30, 36.1, 0, 0, 78.6))

resp_cases$industry <- factor(resp_cases$industry, levels = c("Retail", "General merchandise stores"))

library(ggplot2)
library(plotly)

p <- ggplot(resp_cases, aes(industry, Rates, fill=Year)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), color = "black", width = .7) + 
  scale_fill_manual(values=c("#f08080", "#ffdead", "#bc8f8f")) +
  labs(x = NULL, y = "per 10,000 FTE", fill = NULL, caption = "Source: U.S. Bureau of Labor Statistics") +
  theme_bw() +
  theme(legend.position="top")



ggplotly(p) |> config(displayModeBar = F) |>
  layout(margin = list(l = 50, r = 50, b = 100, t = 50),
         annotations = list(x = 1, y = -0.2, text = "Source: U.S. Bureau of Labor Statistics",
                            xref='paper', yref='paper', showarrow = F, 
                            xanchor='right', yanchor='auto', xshift=0, yshift=0,
                            font = list(size = 9)), 
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y = 1.15))

```
